Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2022-07-19 jsm/speed-presentation/slides.Rmd
minor tweaks
updated slides based on meeting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      fig.lp = '')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
library(ggplot2)
import::from(magrittr, `%>%`)
theme_set(theme_bw())

@@

Michael Trosset,  
Professor of Statistics,  
Indiana University

@@

## Popularity Adjusted Block Model

Let each vertex $i \in [n]$ have $K$ popularity parameters
$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
Then $A \sim \PABM(\{\lambda_{ik}\}_K)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.

**Def** (Noroozi, Rimal, and Pensky, 2020):

$A$ is sampled from a PABM if $P$ can be described as:

1. Let each $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
between communities $k$ and $l$. 
2. Organize popularity parameters as vectors 
$\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
such that $\lambda^{(kl)}_i = \lambda_{k_i l}$ is the popularity parameter 
of the $i$^th^ vertex of community $k$ towards community $l$. 
3. Each block can be decomposed as 
$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.

## Generalized Random Dot Product Graph

**Def** Generalized Random Dot Product Graph  
(Rubin-Delanchy, Cape, Tang, Priebe, 2020)

Let $I_{p,q} = \blockdiag(I_p, -I_q)$ and suppose that $x_1, \ldots, x_n \in \mathbb{R}^{p+q}$ are such that $x_i^\top I_{p,q} x_j \in [0,1]$.  
Then $A \sim \GRDPG_{p, q}(X)$ iff $A \sim \BG(X I_{p,q} X^\top)$, where $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

Adjacency Spectral Embedding (Sussman et al., 2012) estimates $x_1, ..., x_n \in \mathbb{R}^{p+q}$ from $A$:

1. Let $\hat{\Lambda}$ be the diagonal matrix that contains the absolute values of the $p$ most positive and the $q$ most negative eigenvalues.
2. Let $\hat{V}$ be the matrix whose columns are the corresponding eigenvectors.
3. Compute $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.

**Theorem**: $\max\limits_i \|\hat{X}_i - Q_n X_i \| = O_P \Big( \frac{(\log n)^c}{n^{1/2}} \Big)$ 
as $n \to \infty$

## Connecting Block Models to the (G)RDPG Model

@@

:::: column

```{r, out.width = '50%', fig.height = 2.5, fig.width = 2.5}
n1 <- n / 2
n2 <- n / 2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/4
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
P.eigen = eigen(P)
X <- P.eigen$vectors[, 1:2] %*% diag(P.eigen$values[1:2] ** .5)

A.eigen <- eigen(A)
X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
par(mar = rep(1.75, 4))
plot(X, asp = 1, col = z, xlab = NA, ylab = NA, main = 'SBM: Point masses')
```

\vspace*{.5\baselineskip}

```{r, out.width = '50%', fig.height = 2.5, fig.width = 2.5}
par(mar = rep(1.75, 4))
omega <- rbeta(n, 1, 1)
dc.matrix <- omega %*% t(omega)
P.dcbm <- P * dc.matrix
dcbm.eigen <- eigen(P.dcbm)
X.dcbm <- dcbm.eigen$vectors[, 1:2] %*% diag(dcbm.eigen$values[1:2] ** .5)
plot(X.dcbm, asp = 1, col = z, xlab = NA, ylab = NA, main = 'DCBM: Rays')

```

@@

:::: column

\vspace*{0\baselineskip}

```{r, out.width = '100%', fig.height = 4, fig.width = 4}
par(mar = rep(1.75, 4))
Pz <- generate.P.beta(n * 10)
P <- Pz$P
z <- Pz$clustering
X <- embedding(P, 3, 1)
pairs(X, col = z, asp = 1, pch = '.', main = 'PABM: Orthogonal subspaces')
```

@@

## Connecting the PABM to the GRDPG

**Theorem** (KTT): $A \sim \PABM(\{\lambda_{ik}\}_K)$ is equivalent to 
$A \sim \GRDPG_{p, q}(X U)$ with

* $p = K (K + 1) / 2$, $q = K (K - 1) / 2$;
* $U$ is an orthogonal matrix;
* $X \in \mathbb{R}^{n \times K^2}$ is a block diagonal matrix 
composed of popularity vectors with each block corresponding to a community.

$$X = \begin{bmatrix}
\Lambda^{(1)} & \cdots & 0 \\
0 & \ddots & 0 \\
0 & \cdots & \Lambda^{(K)}
\end{bmatrix} 
\in \mathbb{R}^{n \times K^2}$$

$$\Lambda^{(k)} = \begin{bmatrix} 
\lambda^{(k1)} & \cdots & \lambda^{(kK)} 
\end{bmatrix} 
\in \mathbb{R}^{n_k \times K}$$

$$A \sim \PABM(\{\lambda_{ik}\}_K) \text{ iff } A \sim \GRDPG_{p, q}(X U)$$

@@

**Theorem** (KTT): 
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix from OSC. 
Then $\forall$ pairs $(i, j)$ belonging to different communities 
and sparsity factor satisfying $n \rho_n = \omega\big((\log n)^{4c}\big)$,  
$\max_{i, j} \hat{B}_{ij} = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)$.

@@

We compare four algorithms for community detection on randomly generated PABMs:

* Modularity Maximization (Sengupta and Chen) using the Louvain algorithm;
* Orthogonal Spectral Clustering (KTT);
* Sparse Subspace Clustering on the columns of $A$  
(Noorozi, Rimal, Pensky); 
* Sparse Subspace Clustering on the ASE (KTT).

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      fig.lp = '')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
library(ggplot2)
library(ggraph)
import::from(magrittr, `%>%`)
theme_set(theme_bw())

@@

Michael W. Trosset,  
Professor of Statistics,  
Indiana University

@@

## Connecting Block Models to the GRDPG

@@

:::: {.column width='67%'}

```{r, cache = TRUE, fig.width = 6, fig.height = 2, out.width = '100%'}
n <- 64
n1 <- n / 2
n2 <- n / 2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/2
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
X <- embedding(P, 2, 0)
A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)
A.graph <- tidygraph::as_tbl_graph(A, directed = FALSE)

sbm.graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)),
                  show.legend = FALSE) + 
  labs(colour = NULL, 
       title = 'SBM') + 
  # theme_graph() + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

sbm.ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 2], y = Xhat[, 1],
                 colour = factor(z)),
             shape = 4) + 
  geom_point(aes(x = X[, 2], y = X[, 1], 
                 colour = factor(z))) + 
  # coord_fixed() +
  theme_bw() + 
  theme(legend.position = 'none') + 
  labs(x = NULL, y = NULL, colour = NULL, 
       title = 'Point Masses') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(sbm.graph.plot, sbm.ase.plot, nrow = 1)
```

```{r, cache = TRUE, fig.width = 6, fig.height = 2, out.width = '100%'}
n <- 128
n1 <- n / 2
n2 <- n / 2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/2
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
omega <- rbeta(n, 2, 1)
P <- P * (omega %*% t(omega))
X <- embedding(P, 2, 0)
A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)
# Xhat <- -Xhat
A.graph <- tidygraph::as_tbl_graph(A, directed = FALSE)

dcbm.graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)),
                  show.legend = FALSE) + 
  labs(colour = NULL, 
       title = 'DCBM') + 
  # theme_graph() + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

dcbm.ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2],
                 colour = factor(z)),
             shape = 4) + 
  geom_point(aes(x = X[, 1], y = X[, 2], 
                 colour = factor(z))) + 
  # coord_fixed() +
  theme_bw() + 
  theme(legend.position = 'none') + 
  labs(x = NULL, y = NULL, colour = NULL, 
       title = 'Rays') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(dcbm.graph.plot, dcbm.ase.plot, nrow = 1)
```

```{r, cache = TRUE, fig.width = 6, fig.height = 2, out.width = '100%'}
n <- 64
Pz <- generate.P.beta(n)
P <- Pz$P
z <- Pz$clustering
X <- embedding(P, 3, 1)
A <- draw.graph(P)
Xhat <- embedding(A, 3, 1)
Xhat[, 1] <- -Xhat[, 1]
A.graph <- tidygraph::as_tbl_graph(A, directed = FALSE)

pabm.graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)),
                  show.legend = FALSE) + 
  labs(colour = NULL, 
       title = 'PABM') + 
  # theme_graph() + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

pabm.ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2],
                 colour = factor(z)),
             shape = 4) + 
  geom_point(aes(x = X[, 1], y = X[, 2], 
                 colour = factor(z))) + 
  # coord_fixed() +
  theme_bw() + 
  theme(legend.position = 'none') + 
  labs(x = NULL, y = NULL, colour = NULL, 
       title = 'Subspaces (Projected)') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(pabm.graph.plot, pabm.ase.plot, nrow = 1)
```

@@

&nbsp;

&nbsp;

* K-means clustering
* Gaussian mixture models

&nbsp;

* K-means with cosine similarity
* GMM on angles

&nbsp;

&nbsp;

&nbsp;

* ???

@@

@@

**Theorem** (KTT): 
$\forall$ pairs $(i, j)$ belonging to different communities, 
$\max\limits_{i, j} B_{ij} = O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)$.

@@

### 2022-07-19 jsm/speed-presentation/slides.tex
updated slides based on meeting

Michael Trosset,\\
Professor of Statistics,\\
Indiana University
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}{Popularity Adjusted Block Model}
\protect\hypertarget{popularity-adjusted-block-model}{}
\textbf{Def} Popularity Adjusted Block Model (Sengupta and Chen, 2017):

Let each vertex \(i \in [n]\) have \(K\) popularity parameters
\(\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]\). Then
\(A \sim \text{PABM}(\{\lambda_{ik}\}_K)\) if each
\(P_{ij} = \lambda_{i z_j} \lambda_{j z_i}\).

\textbf{Def} (Noroozi, Rimal, and Pensky, 2020):

\(A\) is sampled from a PABM if \(P\) can be described as:

\begin{enumerate}
\tightlist
\item
  Let each \(P^{(kl)}\) denote the \(n_k \times n_l\) matrix of edge
  probabilities between communities \(k\) and \(l\).
\item
  Organize popularity parameters as vectors
  \(\lambda^{(kl)} \in \mathbb{R}^{n_k}\) such that
  \(\lambda^{(kl)}_i = \lambda_{k_i l}\) is the popularity parameter of
  the \(i\)\textsuperscript{th} vertex of community \(k\) towards
  community \(l\).
\item
  Each block can be decomposed as
  \(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\).
\end{enumerate}
\end{frame}

\begin{frame}{Generalized Random Dot Product Graph}
\protect\hypertarget{generalized-random-dot-product-graph}{}
\textbf{Def} Generalized Random Dot Product Graph\\
(Rubin-Delanchy, Cape, Tang, Priebe, 2020)

Let \(I_{p,q} = \text{blockdiag}(I_p, -I_q)\) and suppose that
\(x_1, \ldots, x_n \in \mathbb{R}^{p+q}\) are such that
\(x_i^\top I_{p,q} x_j \in [0,1]\).\\
Then \(A \sim \text{GRDPG}_{p, q}(X)\) iff
\(A \sim \text{BernoulliGraph}(X I_{p,q} X^\top)\), where
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).

Adjacency Spectral Embedding (Sussman et al., 2012) estimates
\(x_1, ..., x_n \in \mathbb{R}^{p+q}\) from \(A\):

\begin{enumerate}
\tightlist
\item
  Let \(\hat{\Lambda}\) be the diagonal matrix that contains the
  absolute values of the \(p\) most positive and the \(q\) most negative
  eigenvalues.
\item
  Let \(\hat{V}\) be the matrix whose columns are the corresponding
  eigenvectors.
\item
  Compute \(\hat{X} = \hat{V} \hat{\Lambda}^{1/2}\).
\end{enumerate}

\textbf{Theorem}:
\(\max\limits_i \|\hat{X}_i - Q_n X_i \| = O_P \Big( \frac{(\log n)^c}{n^{1/2}} \Big)\).
\end{frame}

\begin{frame}{Connecting Block Models to the GRDPG Model}
\protect\hypertarget{connecting-block-models-to-the-grdpg-model}{}
\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{center}\includegraphics[width=0.5\linewidth]{slides_files/figure-beamer/unnamed-chunk-5-1} \end{center}

\vspace*{.5\baselineskip}

\begin{center}\includegraphics[width=0.5\linewidth]{slides_files/figure-beamer/unnamed-chunk-6-1} \end{center}
\end{column}

\begin{column}{0.48\textwidth}
\vspace*{0\baselineskip}

\begin{center}\includegraphics[width=1\linewidth]{slides_files/figure-beamer/unnamed-chunk-7-1} \end{center}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Connecting the PABM to the GRDPG}
\protect\hypertarget{connecting-the-pabm-to-the-grdpg}{}
\textbf{Theorem} (KTT): \(A \sim \text{PABM}(\{\lambda_{ik}\}_K)\) is
equivalent to \(A \sim \text{GRDPG}_{p, q}(X U)\) with

\begin{itemize}
\tightlist
\item
  \(p = K (K + 1) / 2\), \(q = K (K - 1) / 2\);
\item
  \(U\) is an orthogonal matrix;
\item
  \(X \in \mathbb{R}^{n \times K^2}\) is a block diagonal matrix
  composed of popularity vectors with each block corresponding to a
  community.
\end{itemize}

\[X = \begin{bmatrix}
\Lambda^{(1)} & \cdots & 0 \\
0 & \ddots & 0 \\
0 & \cdots & \Lambda^{(K)}
\end{bmatrix} 
\in \mathbb{R}^{n \times K^2}\]

\[\Lambda^{(k)} = \begin{bmatrix} 
\lambda^{(k1)} & \cdots & \lambda^{(kK)} 
\end{bmatrix} 
\in \mathbb{R}^{n_k \times K}\]

\[A \sim \text{PABM}(\{\lambda_{ik}\}_K) \text{ iff } A \sim \text{GRDPG}_{p, q}(X U)\]
\end{frame}

@@

\textbf{Theorem} (KTT): Let \(\hat{B}\) with entries \(\hat{B}_{ij}\) be
the affinity matrix from OSC. Then \(\forall\) pairs \((i, j)\)
belonging to different communities and sparsity factor satisfying
\(n \rho_n = \omega\big((\log n)^{4c}\big)\),\\
\(\max_{i, j} \hat{B}_{ij} = O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)\).

@@

\begin{frame}{Simulation Study}
\protect\hypertarget{simulation-study}{}
We compare four algorithms for community detection on randomly generated
PABMs:

\begin{itemize}
\tightlist
\item
  Modularity Maximization (Sengupta and Chen) using the Louvain
  algorithm;
\item
  Orthogonal Spectral Clustering (KTT);
\item
  Sparse Subspace Clustering on the columns of \(A\)\\
  (Noorozi, Rimal, Pensky);
\item
  Sparse Subspace Clustering on the ASE (KTT).
\end{itemize}

\begin{center}\includegraphics[width=1\linewidth]{slides_files/figure-beamer/unnamed-chunk-8-1} \end{center}
\end{frame}

>>>>

Michael W. Trosset,\\
Professor of Statistics,\\
Indiana University
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}{Connecting Block Models to the GRDPG}
\protect\hypertarget{connecting-block-models-to-the-grdpg}{}
\begin{columns}[T]
\begin{column}{0.67\textwidth}
\begin{center}\includegraphics[width=1\linewidth]{slides_files/figure-beamer/unnamed-chunk-5-1} \end{center}

\begin{center}\includegraphics[width=1\linewidth]{slides_files/figure-beamer/unnamed-chunk-6-1} \end{center}

\begin{center}\includegraphics[width=1\linewidth]{slides_files/figure-beamer/unnamed-chunk-7-1} \end{center}
\end{column}

\begin{column}{0.33\textwidth}
~

~

\begin{itemize}
\tightlist
\item
  K-means clustering
\item
  Gaussian mixture models
\end{itemize}

~

\begin{itemize}
\tightlist
\item
  K-means with cosine similarity
\item
  GMM on angles
\end{itemize}

~

~

~

\begin{itemize}
\tightlist
\item
  ???
\end{itemize}
\end{column}
\end{columns}
\end{frame}

@@

\textbf{Theorem} (KTT): \(\forall\) pairs \((i, j)\) belonging to
different communities,
\(\max\limits_{i, j} B_{ij} = O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)\).

@@

\begin{frame}{Simulation Study}
\protect\hypertarget{simulation-study}{}
\begin{center}\includegraphics[width=1\linewidth]{slides_files/figure-beamer/unnamed-chunk-8-1} \end{center}
\end{frame}
